\pkg{Rgbp}: An R Package for Gaussian, Poisson, and Binomial Hierarchical Modeling

### 2013-08-20 jss/article.tex
paper

\section[introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
\pkg{Rgbp} is an R package for estimating and validating a two-level model, also called conjugate hierarchical generalized linear model \citep{hglm2006, hglm2010}, producing point and interval estiates of the true parameters. The estimation procedure utilizes approximate Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``frequency method checking"). It is found that even in small samples our procedure yields good frequency properties in comparison to other methods such as Maximum Likelihood Estimation (MLE).

@@

\section[Hierarchical Structure]{Hierarchical Structure} \label{hierarchical}
A two-level or multilevel model, also called a conditionally independent hierarchical model \citep{1989}, is a powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

@@

\subsection[Poisson-Gamma]{Poisson-Gamma (``p''=Poisson data)}\label{poissonsubsec}
\code{gbp} is also able to build a Poisson-Gamma hierarchical model. Note that a constant, $1/r$, multiplied to the Gamma distribution below is a scale and a square bracket [, ] below indicates [mean, variance] of a distribution. For notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ as the unbiased estimate of $\lambda_{j}$, where $j=1, \ldots, k$.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}\label{gammaprior}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(r\lambda_{0j})\sim \textrm{Gamma} \left[\lambda_{0j}, ~\frac{\lambda_{0j}}{r} \right],
\end{equation}

@@

where $\lambda^{\ast}_{j} \equiv (1-B_{j})y_{j} + B_{j}\lambda_{0j}$,  $B_{j}\equiv r / (r+n_{j})$, and $y_{j}\equiv z_{j} / n_{j}$, $j=1, \ldots, k$.

@@

To continue with a maximization-based estimation procedure but to steer clear of aforementioned boundary issues we make use of adjustment for density maximization (ADM) \citep{carl1988, 1997, 2011}. For our purposes we approximate the posterior distribution of a shrinkage factor with a Beta distribution, which allows us to finally obtain estimates of the posterior moments, \emph{i.e.}  of $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without the boundary issues that MLE encounters. See \cite{2011}.

@@

There are two kinds of model justification processes; one is a model checking and the other is a method checking. The model checking is for the justification of the hierarchical model on a specific dataset. One possible question is, ``Can this dataset benefit from such a hierarchical modeling?'' \cite{1997} answered this question by using the Negative-Binomial mixture model on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

@@

There are two extreme perspectives that the patient could take in this example. The first being that each hospital is identical in their ability to undertake CABG surgeries, in which case the choice of hospitals is irrelevant and the data could be pooled in order to estimate the death rate. The second, is that the hospitals are totally unrelated in which case the death rate estimate for each hospital would only be a function of the observed data for that particular hospital. These hospitals, however, are all subject to the same rules and regulations enforced by the federal and state governments and that there is likely similarities between hospitals. As such, a hierarchical model seems to be a reasonable approach.

The patient can then presume a state-level (NY) hierarchy governing the true death rates ($\lambda$) of CABG surgery for all the hospitals in NY. This perspective is to view the true death rates of those 31 hospitals as sampled from the state-level population distribution. In this example it is assumed that the true state-level death rate is known to be 0.020.

@@

Hierarchical modeling gives the patient the ability to borrow information across groups. For example, suppose there are two hospitals, whose observed death ratios ($y_{j}$) are 0 and 0.001 and caseloads ($n_{j}$) are 10 and 1000 each. Treating the hospitals as unrelated, the patient may conclude that the former hospital, whose observed death rate is 0, is superior than the latter. Borrowing information from state-level hierarchy seems reasonable for the former hospital because it is hard to judge its true death rate per exposure with just ten caseloads.

@@

From the output provided by \code{gbp} it can be seen that a reasonable choice of a hospital for the patient is hospital 23. Although hospital 8 has a smaller observed mean the fact that hospital 23 has a much higher caseload means that we borrow relatively less information from other hospitals. The result being that the upper bound of the $95\%$ interval is smaller.

% This interval plot provides one more valuable insight, which we could not have noticed. Let's look at the 8th and the 31st hospitals on the graph (or on the outcome for numeric reference). The point estimate of the true death rate per exposure of the 31st hospital is higher than the one of the 8th. But, the upper bound of interval estimate of this 31st hospital is lower than that of the 8th. This interval plot makes the 31st hospital emerge as one of your candidates.

@@

When fitting a model it is always a good idea to question how reliable the estimation procedure is. For example, does our procedure generate interval estimates that have good repeated sampling properties? To answer this question the \code{coverage} function generates pseudo-datasets regarding the estimated $r~(=336.37)$ as a given true value. For reference, we can try any other value of $r$, for example $r=200$, by replacing the code below with `\code{R> pcv <- coverage(p, A.or.r = 200, mean.PriorDist = 0.02, nsim = 1000)}.

@@

The estimated overall average coverage rate is 0.950 from both Rao-Blackwellized and simple unbiased estimates. This result shows that the interval estimates for this particular dataset accurately achieves 95\% confidence under repeated sampling. Note that these estimates depend on the given true value of $r$ and the assumption that the model is true.

@@

One more thing to mention is the accuracy of Rao-Blackwellized unbiased estimates compared to that of simple ones. Taking the first hospital as an example, we can calculate the sample variance of 1,000 Rao-Blackwellized unbiased estimates for the first hospital and that of 1,000 simple ones by typing following code. For reference, \code{raw.resultRB} and \code{raw.resultS}, each row of which is about each hospital, in \code{pcv} contain all the individual simulation results.
\begin{CodeChunk}
\begin{CodeInput}
R> var(pcv$raw.resultRB[1, ])
\end{CodeInput}
\begin{CodeOutput}
 [1] 0.002031225
\end{CodeOutput}
\begin{CodeInput}
R> var(pcv$raw.resultS[1, ])
\end{CodeInput}
\begin{CodeOutput}
 [1] 0.04844745
\end{CodeOutput}
\end{CodeChunk}

It shows that Rao-Blackwellized unbiased estimate is about 24 times more accurate than simple one. Also, as the number of simulation increases (for example 10,000), the zig-zagging blue line will be closer to the red line.
% For reference, it took 10 seconds for MacBook Pro with a 2.3 GHz dual-core Intel i5 CPU to run 10,000 simulations.

@@

All the individual simulation results are saved in \code{pcv$raw.resultRB} and \code{pcv$raw.resultS}, in which each row is about each school.

@@

Conditioning on the true batting average for each player we assume that the at-bats are independent and therefore, $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(45, p_{j}), ~j=1, \ldots, 18$. Our goal is to estimate point and interval estimates of the true batting average, $p_{j}$, for each player. %For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

Suppose, also, that we observed a covariate indicating whether each player is an outfielder or not. It is reasonable to assume that the batting averages for outfielders are different from non-outfielders and therefore we assume different second-level hierarchies for outfielders and for other positions. Within each of which information is shared and regression towards the mean (RTTM) occurs. \code{gbp} provides a way to incorporate such information (position) seamlessly into the second-level hierarchy.

@@

We see that the two prior means distinguishing outfielders from other positions are significantly different (p-value = 0.038), supporting our initial belief. The positive sign of $\hat{\beta}_{1}$ indicates that the mean batting average for all the outfielders tends to be higher than that for those in the other positions.

@@

Then, how much can we trust these interval estimates? The following frequency method checking that regards the estimated values, $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values will answer it. For reference, if we want to try different true values, the appropriate code will be \code{coverage(b, A.or.r = 100, reg.coef = c(-1, 0.2), nsim = 1000)} instead of below one.

>>>>

\section[introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
\pkg{Rgbp} is an R package for estimating and validating a two-level model, also called conjugate hierarchical generalized linear model \citep{hglm2006, hglm2010} or called a conditionally independent hierarchical model \citep{1989}, producing point and interval estimates of the true parameters. The estimation procedure utilizes approximate Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``frequency method checking"). It is found that even in small samples our procedure yields good frequency properties in comparison to other methods such as Maximum Likelihood Estimation (MLE).

@@

\section[Hierarchical Structure]{Hierarchical Structure} \label{hierarchical}
A two-level or multilevel model is a powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

@@

\subsection[Poisson-Gamma]{Poisson-Gamma (``p''=Poisson data)}\label{poissonsubsec}
\code{gbp} is also able to build a Poisson-Gamma hierarchical model. Note that a constant, $1/r$, multiplied to the Gamma distribution below is a scale and a square bracket [, ] below indicates [mean, variance] of a distribution. For notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ as the unbiased estimate of $\lambda_{j}$ for the $j$-th group.
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}\label{gammaprior}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(r\lambda_{0j})\sim \textrm{Gamma} \left[\lambda_{0j}, ~\frac{\lambda_{0j}}{r} \right],
\end{equation}

@@

where $\lambda^{\ast}_{j} \equiv (1-B_{j})y_{j} + B_{j}\lambda_{0j}$ and   $B_{j}\equiv r / (r+n_{j})$ for $j=1, \ldots, k$.

@@

To continue with a maximization-based estimation procedure but to steer clear of aforementioned boundary issues we make use of adjustment for density maximization (ADM) \citep{carl1988, 1997, 2011}. For our purposes we approximate the posterior distribution of a shrinkage factor with a Beta distribution, which allows us to finally obtain estimates of the posterior moments, \emph{i.e.}  of $E(B_{j}\vert\textbf{y})$ and $Var(B_{j}\vert\textbf{y})$, where the notation $\textbf{y}$ represents observed data, without the boundary issues that MLE encounters. See \cite{2011}.

@@

There are two kinds of model justification processes; one is a model checking and the other is a method checking. The model checking is for the justification of the hierarchical model on a specific dataset. One possible question is, ``Can this dataset benefit from such a hierarchical modeling?'' \cite{modelchecking1996} answered this question by using the Negative-Binomial mixture model on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

@@

There are two extreme perspectives that the patient could take in this example. The first is that each hospital is identical in their ability to undertake CABG surgeries, in which case the choice of hospitals is irrelevant and the data could be pooled in order to estimate the common death rate for all hospitals (complete pooling). The second, is that the hospitals are totally unrelated in which case the death rate estimate for each hospital would only be a function of the observed data for that particular hospital (no pooling).

However, in reality, these hospitals are neither completely the same nor different. For example, doctor' ability and experience for dealing with such a surgery vary according to each hospital. At the same time, these hospitals are all subject to the same rules and regulations enforced by the federal and state governments. The hierarchical model provides a happy medium accounting for such a difference and similarity altogether (partial pooling).

The patient can then presume a state-level (NY) hierarchy governing the true death rates ($\lambda$) of CABG surgery for all the hospitals in NY. This perspective is to view the true death rates of those 31 hospitals as sampled from the state-level population distribution. In this example it is assumed that the mean of such state-level death rate distribution is known to be 0.020 ($\lambda_{0}$).

@@

Intuitively, switching ranks for these two hospitals makes sense. To be specific, their observed death ratios ($y_{j}$, $j=8, 23$) are 0.014 and 0.015 and caseloads ($n_{j}$, $j=8, 23$) are 295 and 602 each. Considering solely the observed death ratios may lead to an unfair comparison because the latter hospital handled more than twice caseloads. \pkg{Rgbp} accounts for this caseload difference, making the estimate of true death rate for the former hospital shrink toward the state-level mean ($\lambda_{0}$=0.02) much harder than that for the latter hospital. This is what happened to the two left-most hospitals in the first plot.

@@

This interval plot confirms that the 23rd hospital has the lowest upper bound for the true death rate, not to mention the lowest point estimate. Also it reveals that the 31st hospital may be the second safest because its upper bound of 95\% interval estimate for the true death rate is lower than the 8th hospital's one.

%From the output provided by \code{gbp} it can be seen that a reasonable choice of a hospital for the patient is hospital 23. Although hospital 8 has a smaller observed mean the fact that hospital 23 has a much higher caseload means that we borrow relatively less information from other hospitals. The result being that the upper bound of the $95\%$ interval is smaller.

%provides one more point, which we could not have noticed. Let's look at the 8th and the 31st hospitals on the graph (or on the outcome for numeric reference). The point estimate of the true death rate per exposure of the 31st hospital is higher than the one of the 8th. But, the upper bound of interval estimate of this 31st hospital is lower than that of the 8th. This interval plot makes the 31st hospital emerge as one of your candidates.

@@

Then, how reliable are these intervals? Does our procedure generate interval estimates that have good repeated sampling properties? To answer this question the \code{coverage} function generates pseudo-datasets regarding the estimated $r~(=336.37)$ as a given true value. For reference, we can try any other value of $r$, for example $r=200$, by replacing the code below with `\code{R> pcv <- coverage(p, A.or.r = 200, mean.PriorDist = 0.02, nsim = 1000)}.

@@

The estimated overall average coverage rate across all the hospitals is 0.950 from both Rao-Blackwellized and simple unbiased estimates. This result shows that the interval estimates for this particular dataset accurately achieves 95\% confidence under repeated sampling. Note that these estimates depend on the given true value of $r$ and the assumption that the model is true.

@@

One more thing to mention is the accuracy of Rao-Blackwellized unbiased estimates compared to that of simple ones. Taking the first hospital as an example, we can calculate the sample variance of 1,000 Rao-Blackwellized unbiased estimates for the first hospital and that of 1,000 simple ones by typing following code. For reference, \code{raw.resultRB} and \code{raw.resultS}, each row of which is about each hospital, in \code{pcv} contain all the individual simulation results.
\begin{CodeChunk}
\begin{CodeInput}
R> var(pcv$raw.resultRB[1, ]) / 1000
\end{CodeInput}
\begin{CodeOutput}
 [1] 0.000002031225
\end{CodeOutput}
\begin{CodeInput}
R> var(pcv$raw.resultS[1, ]) / 1000
\end{CodeInput}
\begin{CodeOutput}
 [1] 0.00004844745
\end{CodeOutput}
\end{CodeChunk}

It shows that Rao-Blackwellized unbiased estimate is about 24 times more accurate than simple one. Also, as the number of simulation increases, the zig-zagging blue line is expected to be closer to the red line.
% For reference, it took 10 seconds for MacBook Pro with a 2.3 GHz dual-core Intel i5 CPU to run 10,000 simulations.

@@

All the individual simulation results are saved in \code{gcv$raw.resultRB} and \code{gcv$raw.resultS}, in which each row is about each school.

@@

Conditioning on the true batting average for each player we assume that the at-bats are independent and therefore, $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(45, p_{j}), ~j=1, \ldots, 18$. Our goal is to estimate point and interval estimates of the true batting average, $p_{j}$, for each player, considering additional information on whether the player is an outfielder or not. \code{gbp} provides a way to incorporate such information seamlessly into the second-level hierarchy so that information is shared and regression towards the mean (RTTM) occurs within outfielders and non-outfielders separately. %For reference, we can also assume the Normal distribution using $y_{j}=z_{j} / n_{j}$ and $V_{j}=\bar{y}(1-\bar{y})/n$, where $\bar{y}=\sum_{j} z_{j} / \sum_{j} n_{j}$, $j=1,\ldots, 18$.

@@

We see that the two prior means distinguishing outfielders from other positions are significantly different (p-value = 0.038). Also, the positive sign of $\hat{\beta}_{1}$ indicates that the population mean batting average for all the outfielders tends to be higher than that for those in the other positions (estimated odds ratio = exp(0.389)=1.48).

@@

Then, how much can we trust these interval estimates? The following frequency method checking that regards the estimated values, $\hat{r}$ (=112.95) and $\hat{\beta}~(=(-1.194, ~0.389)^{T})$, as true values will answer it. For reference, if we want to try different true values for example 100 for $r$ and (-1, 0.2) for $(\beta_{0}, \beta_{1})$, the appropriate code will be \code{coverage(b, A.or.r = 100, reg.coef = c(-1, 0.2), nsim = 1000)} instead of below one.
